Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-03-18 young_pp_discs.tex
updataed figs, minor text changes

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt*{Joos2012,Hayfield2011,Machida2011}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing envelope material. \smallskip

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, L1521F, a similar dense core, contains a central object which appears to be a low mass protostar and disc \citep{Bourke2006}, whilst having between half and three quarters of the mass of L1544. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Figure \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing gas (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 10 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT6.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities2.eps}
 \caption{Column densities of OCS, H$_2$CO,HCO$^+$ and CO used in the disc model. Figure adapted from I2011.}
 \label{Chemistry} 
\end{figure*}

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (figure \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 1/100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history. In the dense conditions of the protoplanetary disc could be encouraging the dust grains to agglomerate and increase the size of the ice mantle through freeze out of volatiles. This could lead to changes in the dust opacity, however we have not attempted to include these effects in our model. The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However the results described here do not depend on the details of the dust opacity and would be qualitatively similar if other opacities were used.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011 who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamics simulation included Lagrangian tracer fluid elements in addition to solving the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au in x, y, and z to reveal the morphology of the disk as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 20\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and one which traces the entirety of the disc (C$^{17}$O) (see figure \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \cite{Caselli2012}). As discussed in the result section \ref{sec:model_results}, approximations have been made based on values measured toward similar objects. The molecular data used is from the Leiden Atomic and Molecular DAtabase (LAMDA) (Sch\"oier et al. 2005; http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \cite{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay triangulation means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (figure \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. In order to minimise the artifacts in the output images, resulting from the grid construction, the average of ten separate runs was taken (Figure \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime2.eps}%Lime_grid3.eps
 \caption{A plot of the points selected by the griding process and the paths down which photons can propagate for points in the central r,$\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical coordinates, linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\theta$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum.eps}
 \caption{{\bf Left:}A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite grid-ding.}
 \label{averages}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo2.eps}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions.}
 \label{points}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way).For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the $^{17}$O/$^{16}$O elemental abundance. These estimates for the abundances in the envelope are simplistic, however, as seen in section \ref{sec:alma_predictions}, the envelope contribution to the line is very narrow, and can be spectrally disentangled from the disc contribution.\smallskip

@@

%We considered C$^{17}$O 3$\rightarrow$2 at 337.1$\,$GHz with an upper energy level of $E_u$=32.3K, HCO$^+$ 3$\rightarrow$2 at 267$\,$GHz and $E_u$=25.7K, OCS 26$\rightarrow$25 at 316$\,$GHz and $E_u$=205K and H$_2$CO 4$_{04}\rightarrow$3$_{03}$ at 290$\,$GHz and $E_u$=34.9K.\smallskip

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ to edge on, we will consider the 30$^\circ$ case as a 'typical' inclination with the 15$^\circ$ inclination discussed in section \ref{sec:alma_predictions}, the other inclinations simulated are shown in appendix \ref{sec:other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity, intensity weighted velocity and channel maps (figures \ref{sim_all}, \ref{h2co_chanmap} and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope. This can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structure in different regions of the disc is highlighted by looking at different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are integrated over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (figure \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}) it can be used to view rotation of the central part of the disc without risk of contamination from emission from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that it is not visible throughout all the regions in the disc where it is found, the line is only seen at radii <16$\,au$ whereas the molecules have a significant abundance out to $\sim$30$\,$au. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26 - 25), as shown in figure \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, because of the presence of the protostar and other hot material within 2$\,$au of the protostar, not considered in our simple model, we expect absorption to also be observed toward the central region of the disc.\smallskip

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$au of the disc, which can be seen in the integrated intensity map and its rotation is detected out to larger radii than is possible with the OCS line (figure \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. Unlike the OCS however, the H$_2$CO extends out to large enough radii to be present in the voids between the outer parts spiral arms. In these regions, the H$_2$CO line can be seen in emission.\smallskip

@@

HCO$^+$ in unique in the molecules simulated in I2011 in that it traces only the outer regions of the disc, and so can be used to look at the extended velocity and physical structure. In figure \ref{sim_all} we can see that the HCO$^+$ line shows the outer edges of the spiral arms in absorption, but it also shows emission from the diffuse gas further out in the disc. This allows the velocity structure of the disc to be measured out to larger radii and allows constraints to be placed on the rotation of the disc out to larger radii than with other molecules.

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate their observation with the completed ALMA in the 26th most extended configuration (out of a total of 28), with the longest baseline of 14.4km and giving beam sizes between 0.02 and 0.03 arcseconds at the selected frequencies, giving a beam size close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{80mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of $\sim$400$\,$ms$^{-1}$. As calculated by the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06GHz & 32.3$\,$K & 0.0270$\,$mJy$\,$beam$^{-1}$ \\
      OCS & 26$\rightarrow$25 & 316.15GHz & 25.7$\,$K & 0.0275$\,$mJy$\,$beam$^{-1}$ \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56GHz & 205$\,$K & 0.0176$\,$mJy$\,$beam$^{-1}$ \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62GHz & 34.9$\,$K & 0.0228$\,$mJy$\,$beam$^{-1}$ \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

>>>>

The formation and early evolution of protoplanetary discs around solar-type and low-mass protostars has little observational support, despite of the fast growing list of theoretical models on the dynamical evolution of star forming dense cores (e.g. \citealt{Krasnopolsky2011,Machida2011,Braiding2012,Joos2013}). The reason for this is that young protostars are surrounded by thick envelopes and power energetic outflows. Thus, observations of the young discs, predicted to have sizes of about 100\,au and masses as large as 10\% the original core mass (e.g. \citealt*{Joos2012,Hayfield2011,Machida2011}), are challenging. The use of sensitive interferometers is needed to achieve high angular resolution and spatially/spectrally disentangle the various disc, envelope and outflow components as well to filter out the extended emission tracing the envelope material. \smallskip

@@

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, it is very similar in structure to L1521F,  another dense core in Taurus which hosts a low mass protostar and possibly a disc \citep{Bourke2006}. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Fig. \ref{fig:l1544_model} shows the physical parameters of the core model. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/L1544model_used_legend_small2.eps}
 \caption{The spherically symmetric model of the pre-stellar core L1544 used as the envelope of the young protoplanetary disc in the hybrid model. Showing gas (green) and dust (red) temperature in kelvin, log number density (blue) in cm$^{-3}$ and inward velocity / 10 (cyan) in m$\,$s$^{-1}$. Adapted from KC2010 and Keto et al. (2013, in preparation).}
 \label{fig:l1544_model}
\end{figure}

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/rhoT5.eps}
 \caption{{\bf Left:} A 3D plot of log number density (cm$^{-3}$) showing the spiral structure in the xy plane and scale height of the disc. {\bf Right:} The 3D temperature (K) structure of the disc; regions cooler than 40$\,$K are not shown in 3D, highlighting the narrow central region containing hot material.}
 \label{rhoT} 
\end{figure*}

\begin{figure*}
 \includegraphics[width=168mm]{Figures/model/columnDensities3.eps}
 \caption{Column density distributions of OCS, H$_2$CO, HCO$^+$ and CO across the disc, as calculated by I2011. Note the different zones traced by the various species, with OCS mostly tracing the inner regions and HCO$^+$ probing the gas at larger radii.}
 \label{Chemistry} 
\end{figure*}

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.~\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y, and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to solving the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y, and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer is taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/).

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \citealt{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation\footnote{For three dimensions, Delaunay triangulation means that if four points are connected into a tetrahedron, the sphere circumscribing these four points contains no other points. It can be shown that this connection is unique for a given set of points.} and it is down these paths that photon propagation is restricted (fig. \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 in 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. The average of ten separate runs was taken to minimise the artifacts in the output images, resulting from the grid construction (Fig. \ref{averages}).

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime3.eps}
 \caption{A plot of the points selected by the gridding process and the paths down which photons can propagate for points in the central r, $\theta$ plane. The points are colour coded by the density distribution (in m$^{-3}$, as used in LIME) and are more concentrated at small radii and in the most dense regions. The circular high density region in the centre is the disc model and is 128$\,$au in diameter.}
 \label{grid}
\end{figure}

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their density and molecular density compared against a reference point in order to decide if the point is to be used in the grid or not. Candidate grid points are selected at random in cylindrical coordinates, linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, otherwise another r, $\theta$, z co-ordinate is selected and this becomes the candidate point. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. \smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/continuum2.eps}
 \caption{{\bf Left:} A 300$\,$GHz continuum image of the model, created from the average of ten LIME runs. {\bf Right:} the output of a single radiative transfer simulation, with artifacts due to finite gridding.}
 \label{averages}
\end{figure}

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/lime_points_rz_histo3.eps}
 \caption{A 2D histogram of the point distribution throughout the model. The disc and envelope can be seen as two separate entities which have to be sampled using different point distributions.}
 \label{points}
\end{figure}

@@

Simulations are limited to molecules for which we have abundances in both the disc and envelope, and also have calculated Einstein and collisional coefficients. In the envelope, the HCO$^+$ abundance profile follows the H$_2$O profile of L1544 \citep{Caselli2012}, scaled so the maximum is 1$\times$10$^{-8}$ (assuming, for simplicity, that photodissociation and freeze-out affect these two species in a similar way). For H$_2$CO, a step profile was considered, with a fractional abundance of 1.5$\times$10$^{-8}$ at radii greater than 0.04$\,$pc and 1.5$\times$10$^{-9}$ inside it \citep{Young2004}. For OCS we adopted a constant abundance of 1.9$\times$10$^{-9}$ \citep{Ren2011}. The C$^{17}$O abundance profile in the envelope follows the CO profile calculated for L1544 by KC2010, scaled by the $^{17}$O/$^{16}$O elemental abundance. These estimates for the abundances in the envelope are simplistic, however, as seen in section \ref{sec:alma_predictions}, the envelope contribution to the line is very limited in velocity and can be spectrally disentangled from the disc contribution.\smallskip

@@

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ relative to the edge on case.  In the following, we will show the 30$^\circ$ case, while the 15$^\circ$ is discussed in section \ref{sec:alma_predictions}. Other inclinations are shown in appendix \ref{sec:other_inc}. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and  intensity weighted velocity maps (fig.s \ref{sim_all}, and \ref{mom0_maps}) were created by integrating between -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, which can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

\begin{figure*}
 \includegraphics[width=150mm]{Figures/sim/imageALL_30deg_all2.eps}
 \caption{{\bf Left:} Continuum subtracted integrated intensity maps. {\bf Right:} Intensity weighted velocity maps. The spiral structure in different regions of the disc is highlighted by looking at different molecular lines. From top to bottom the lines displayed are: OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2. All maps are integrated over -12.5 to -0.5 and 0.5 to 12.5 km$\,$s$^{-1}$ in order to avoid the envelope contribution.}
 \label{sim_all}
\end{figure*}

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (fig. \ref{sim_all}). As OCS is not seen in outflows (e.g. \citealt{Stanke2007}, \citealt{VDTak2003}), it can be used to measure the rotation of the central part of the disc without contamination from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that they trace radii $<$16$\,$au, even smaller than the extent of the OCS column density distribution shown in fig.\ref{Chemistry}. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26$\rightarrow$25), as shown in fig. \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, this may be an artefact of our simplistic treatment of the inner hole, given that we neglect the presence of the protostar and other hot material within the central 2$\,$au. Thus, we expect absorption to be also observed toward the central region of the disc.\smallskip

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$au of the disc, which can be seen in the integrated intensity map, and its rotation is detected out to larger radii compared to the OCS line (fig. \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. However, unlike the OCS, the H$_2$CO extends out to large enough radii to trace the voids between the outer spiral arms. In these regions, the H$_2$CO line can be seen in emission.\smallskip

@@

The HCO$^+$ line only traces the outer regions of the disc. In fig. \ref{sim_all} we can see that the HCO$^+$ line shows the outer edges of the spiral arms in absorption, but it also shows emission from the diffuse gas further out in the disc. This allows the velocity structure of the disc to be measured out to larger radii, allowing constraints to be placed on the rotation of the disc out to larger radii than with other molecules.

@@

In order to make predictions on the observability of the synthetic brightness maps, the Common Astronomy Software Applications package (CASA) was used to simulate observations with the completed ALMA in the 26th most extended configuration (out of a total of 28). The longest baseline of this configuration is 14.4\,km and the beam sizes range between 0.02 and 0.03 arcseconds at the selected frequencies. This angular resolution is close to the size of the spiral structure in the disc. As the weakest lines studied in the model are of the order of 0.1-0.2$\,$mJy$\,$beam$^{-1}$ at this resolution, Band 7 sensitivities of the order of 0.02$\,$mJy$\,$beam$^{-1}$ are required. This implies around 6 hours of integration time. The exact sensitivities are given in Table \ref{sigmas}.
\begin{table*}
  \centering
  \begin{minipage}{90mm}
    \caption{Line sensitivities obtained with ALMA after 6 hours integration time with velocity resolution of 400$\,$m\,s$^{-1}$, using the ALMA on-line sensitivity calculator.}
    \label{sigmas}
    \begin{tabular}{c||c|c|c|c}
      \hline
      Species & Transition & Frequency & Upper energy level & Sensitivity\\
                      &                    & (GHz)            & (K)                            & (mJy beam$^{-1}$) \\
      \hline
      C$^{17}$O & 3$\rightarrow$2 & 337.06 & 32.3 & 0.0270 \\
      OCS & 26$\rightarrow$25 & 316.15 & 205 & 0.0275 \\
      HCO$^+$ & 3$\rightarrow$2 & 267.56 & 25.7 & 0.0176 \\
      H$_2$CO & 4$_{04}\rightarrow$3$_{03}$ &  290.62 & 34.9 & 0.0228 \\
      \hline
    \end{tabular}
  \end{minipage}
\end{table*}
In order to clean the images which featured mainly absorption, the sky model used was the output of the LIME simulations with the continuum subtracted and then the spectrum inverted. The simulated images were cleaned down to a 3$\sigma$ level with $\sigma$ (the sensitivity) given in Table \ref{sigmas}.\smallskip

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/casa_cont_300GHz_invert2.eps}
